An Effective Model of HL-60 Differentiation

### 2017-04-19 Paper_v2.tex
Working on discussion

The PU.1 and AP1 proteins were important regulators of the ATRA-induced signal integration and phenotypic programs (Fig. \ref{fig:model-response-ko}).
We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
All simulations were conducted using the best fit parameter set.
The first four modes described $>$99\% of the gene knockout variance,
with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
However, this analysis did not give insight into which upstream components of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.

@@

We hypothesized that Vav1 is a member of an ATRA-inducible complex which propels sustained MAPK activation, arrest and differentiation.
Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
This increased interaction facilitates phosphorylation and activation of Raf by pre-bound Akt and/or CK2 at S621 or perhaps S259.
Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors,
e.g. Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
We tested this working hypothesis using mathematical modeling.
The model recapitulated both ATRA time-course data as well as the GW5074 inhibitor effects.
This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
Further, analysis of the Raf-Vav1 model identified bistability in ppERK levels.
Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values. The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
sustained signaling mode (high ppERK) even in the absence of a ATRA signal.
This locked-in property could give rise to an ATRA-induction memory.
We validated the treatment memory property predicted by the Raf-Vav1 circuit experimentally using ATRA-washout experiments.
ERK phosphorylation levels remained high for more then 96 hr after ATRA was removed.
Previous studies demonstrated that HL-60 cells possessed an inheritable memory of ATRA stimulus \cite{YEN1984}.
Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels
over the entire ensemble.
CD38 expression correlated with the ppERK, suggesting its involvement in the signaling complex.
Our computational and experimental results showed that positive feedback,
through ERK-dependent Raf expression, could sustain MAPK signaling through many division cycles.
Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.

Several engineered, or naturally occurring systems involved in cell fate decisions incorporate positive feedback and bistability \cite{Ferrell:2002aa}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the differentiation circuit presented here; ATRA drives signalsome which activates MAPK, cell-cycle arrest, differentiation and signalsome.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, they share similar cell fate decision architectures.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while positive feedback is sometimes not desirable in man made systems, it may be at the core of a diverse variety of
cell fate programs and other networks important to human health.

Model performance was impressive given its limited size. However, there were several issues to explore further.
First, there was likely missing connectivity in the effective differentiation circuit.
Decreasing BLR1 expression with simultaneously sustained cRaf-pS261 activation was not captured by the current network architecture.
This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
We could model this by separating signalsome formation into an inactive precursor pool that is transformed to a long-lived activated signalsome by MAPK activation.
We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules might have different properties which could influence model identification or performance.
For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
The form of the transfer function could also be explored. We choose hill-like functions because of their
prominence in the systems and synthetic biology community. However, many other transfer functions are possible.

>>>>

The PU.1 and AP1 proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
All simulations were conducted using the best fit parameter set.
The first four modes described $>$99\% of the gene knockout variance,
with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent Gfi1 and C/EBPa (Fig. \ref{fig:model-response-ko}A).
To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
The most sensitive interactions for the PU.1 protein involved the C/EBPa-dependent regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
On the other other, the most sensitive connections for AP1 involved the C/EBPa-dependent regulation of p21 expression, and the mutual activation of PU.1 and AP1 expression.
Taken together, these results suggested that the PU.1 and AP1 proteins acted as important self-reinforcing regulators for both the signal integration and phenotype modules.
The analysis suggested that the PU.1 signaling axis promoted the formation of the neutrophil NADPH oxidase (through p47Phox), while AP1 was responsible for cell cycle arrest (through p21).
However, this analysis did not give insight into which upstream components of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

Analysis of the signal integration and phenotype modules suggested that PU.1 and AP1 regulated distinct phenotypic axes following ATRA exposure.
In particular, deletion of PU.1 and AP1 was predicted to lead to largest shift in the system state.

@@

We hypothesized that Vav1 is a member of an ATRA-inducible complex which propels sustained MAPK activation, arrest and differentiation.
Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
This increased interaction facilitates phosphorylation and activation of Raf by pre-bound Akt and/or CK2 at S621 or perhaps S259.
Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors
e.g., Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
We tested this working hypothesis using mathematical modeling.
The model recapitulated both ATRA time-course data as well as the GW5074 inhibitor effects.
This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
Further, analysis of the Raf-Vav1 model identified bistability in ppERK levels.
Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values. The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
sustained signaling mode (high ppERK) even in the absence of a ATRA signal.
This locked-in property could give rise to an ATRA-induction memory.
We validated the treatment memory property predicted by the Raf-Vav1 circuit experimentally using ATRA-washout experiments.
ERK phosphorylation levels remained high for more then 96 hr after ATRA was removed.
Previous studies demonstrated that HL-60 cells possessed an inheritable memory of ATRA stimulus \cite{YEN1984}.
Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on ppERK levels
over the entire ensemble.
CD38 expression correlated with the ppERK, suggesting its involvement in the signaling complex.
Our computational and experimental results showed that positive feedback,
through ERK-dependent Raf expression, could sustain MAPK signaling through many division cycles.
Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.

Several engineered, or naturally occurring systems involved in cell fate decisions incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
Reinforcing feedback and bistability has also been implicated in hematopoietic stem cell fate determination.
Laslo et al showed that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBPa), encodes a bistable switch that
results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBPa expression \cite{Laslo2006}.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while positive feedback is often undesirable in human engineered systems, it may be at the core of a diverse variety of
cell fate programs and other networks important to human health.

There were several issues that can be explored further with the effective ATRA differentiation model.
First, there was likely missing connectivity in the effective differentiation circuit.
Decreasing BLR1 expression with simultaneously sustained cRaf-pS261 activation was not captured by the current network architecture.
This suggested that signalsome, once activated, had a long lifetime as decreased BLR1 expression did not impact cRaf-pS261 abundance.
We could model this by separating signalsome formation into an inactive precursor pool that is transformed to a long-lived activated signalsome by MAPK activation.
We should also explore adding additional downstream biological modules to this skeleton model, for example the upregulation of reactive oxygen markers such as p47Phox or
cell cycle arrest components to capture the switch from an actively proliferating population to a population in G0-arrest.
Next, the choice of max/min integration rules or the particular form of the transfer functions could also be explored.
Integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules might have different properties which could influence model identification or performance.
For example, a mean integration rule would be differentiable, allowing derivative-based optimization approaches to be used.
The form of the transfer function could also be explored. We choose hill-like functions because of their
prominence in the systems and synthetic biology community. However, many other transfer functions are possible.
